---
title: "Professional sanctions in AI hallucination cases"
format:
  html:
    embed-resources: true
    toc: true
---

Group members: Ishita Jain, Ahlam Negash, Abigail Minin

GitHub repository: [View our GitHub repository](https://github.com/ishitajain21/stat-learning-group-analysis)

Partner 1: Ishita Jain — research question, exploratory analysis, decision tree.

Partner 2: random forest, comparison, conclusions (name to be confirmed).

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5)
if (dir.exists("../.hw5-R-library")) .libPaths(c(normalizePath("../.hw5-R-library"), .libPaths()))
library(tidyverse)
library(rpart)
library(rpart.plot)
```

## Partner 1: Research question

Can party type, jurisdiction, AI-tool identification, and primary legal field
classify whether a documented AI hallucination case records a professional sanction?
The response is `Professional Sanction`, with Yes as the positive class.
This replaces the starter's question about counts over time to meet the
assignment's classification requirement.

The data come from Damien Charlotin's AI Hallucination Cases Database.
Each row represents a documented case, not a randomly sampled use of AI.
We predict the recorded label; No does not rule out later sanctions.

## Partner 1: Exploratory analysis

```{r}
cases <- read_csv("data/hallucination_cases.csv")
knitr::kable(cases |> count(`Professional Sanction`, name = "Cases"))
model_data <- cases |>
  filter(`Professional Sanction` %in% c("No", "Yes")) |>
  transmute(
    sanction = factor(`Professional Sanction`, levels = c("No", "Yes")),
    party = case_when(
      str_detect(str_to_lower(coalesce(`Party(ies)`, "")), "lawyer|prosecutor|defender") ~ "Lawyer involved",
      str_detect(str_to_lower(coalesce(`Party(ies)`, "")), "pro se") ~ "Pro se litigant",
      is.na(`Party(ies)`) ~ "Unknown",
      TRUE ~ "Other party"),
    jurisdiction = case_when(`State(s)` == "USA" ~ "USA",
                             is.na(`State(s)`) ~ "Unknown", TRUE ~ "Outside USA"),
    tool = case_when(str_to_lower(`AI Tool`) == "chatgpt" ~ "ChatGPT",
                    str_to_lower(`AI Tool`) %in% c("implied", "unidentified") | is.na(`AI Tool`) ~ "Unspecified or implied",
                    TRUE ~ "Other named tool"),
    legal_field = coalesce(`Legal Field Primary`, "Unknown"))
sanction_rate <- mean(model_data$sanction == "Yes")
ggplot(model_data, aes(sanction)) + geom_bar(fill = "steelblue") +
  labs(x = "Recorded professional sanction", y = "Cases",
       title = "Professional sanctions are uncommon") + theme_minimal()
party_summary <- model_data |> group_by(party) |>
  summarise(Cases = n(), Yes = sum(sanction == "Yes"),
            Sanction_rate = mean(sanction == "Yes"), .groups = "drop")
knitr::kable(party_summary, digits = 3)
ggplot(party_summary, aes(reorder(party, Sanction_rate), Sanction_rate)) +
  geom_col(fill = "darkorange") + coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Party category", y = "Proportion with sanction",
       title = "Recorded sanction rates by party type") + theme_minimal()
```

There are `r nrow(cases)` cases; `r nrow(cases) - nrow(model_data)` are
excluded for missing or unrecognized responses. Among retained cases,
`r round(100 * sanction_rate, 2)`% record a sanction. The imbalanced response
means high accuracy can conceal poor detection of Yes cases.
The party chart compares rates rather than counts to account for category
sizes. Lawyer involvement may be informative because professional discipline
applies to legal professionals. These associations do not establish causation.

## Shared modeling setup

```{r}
set.seed(154)
train_id <- sample(seq_len(nrow(model_data)), round(0.7 * nrow(model_data)))
train <- model_data[train_id, ]
test <- model_data[-train_id, ]
# Learn common fields from training data only; group unseen test levels.
common_fields <- train |> count(legal_field, sort = TRUE) |>
  slice_head(n = 8) |> pull(legal_field)
prepare_predictors <- function(d) {
  d |> mutate(
    legal_field = factor(ifelse(legal_field %in% common_fields, legal_field, "Other"),
                         levels = c(common_fields, "Other")),
    party = factor(party, levels = c("Lawyer involved", "Pro se litigant", "Other party", "Unknown")),
    jurisdiction = factor(jurisdiction, levels = c("USA", "Outside USA", "Unknown")),
    tool = factor(tool, levels = c("ChatGPT", "Unspecified or implied", "Other named tool")))
}
train <- prepare_predictors(train)
test <- prepare_predictors(test)
model_formula <- sanction ~ party + jurisdiction + tool + legal_field
knitr::kable(data.frame(Set = c("Training", "Test"),
                       Cases = c(nrow(train), nrow(test)),
                       Yes_rate = c(mean(train$sanction == "Yes"), mean(test$sanction == "Yes"))), digits = 3)
```

The seed-154 70/30 split gives `r nrow(train)` training and `r nrow(test)`
test cases. Partner 2 must reuse these processed sets and `model_formula`.
Outcome text, monetary penalties, and case narratives are excluded because
they could reveal the response; names and source links are also excluded.
This is a retrospective random holdout, not a forecast of future cases.

## Partner 2: Random forest

Pending Partner 2. Fit a random forest with `model_formula`, `train`, and
`test` above. Use Yes as the positive class and a 0.5 cutoff for the initial
comparison. Evaluate with `classification_metrics()`.

## Partner 2: Model comparison and conclusions

Pending Partner 2 and shared review. Compare with `tree_metrics` and the
always-No baseline using accuracy, sensitivity, specificity, balanced
accuracy, and interpretability. Discuss class imbalance and data limitations.
