---
title: "AI Hallucination Cases"
format: html
---

Group members: Ishita Jain, Ahlam Negash, Abigail Minin
Research Question: Has the number of documented AI hallucination court cases increased over time?

GitHub repository: [View our GitHub repository](https://github.com/ishitajain21/stat-learning-group-analysis)

The data come from the AI Hallucination Cases Database maintained by Damien Charlotin. This file is the shared starter for the group analysis. Develop one research question together, then add the exploratory analysis and model here. Before editing, pull. After a checkpoint, pull again, then push. Avoid editing the same section at the same time.

```{r}
library(tidyverse)

cases <- read_csv("data/hallucination_cases.csv")
glimpse(cases)
```

```{r}
#Data Visualization 1 
ggplot(cases, aes(x = `Party(ies)`)) +
  geom_bar() +
  coord_flip() +
  labs(x = "Party", y = "Number of cases", title = "Cases by party")
```

```{r}
#Data Visualization 2
ggplot(cases, aes(x = `Legal Field Primary`)) +
  geom_bar() +
  coord_flip() +
  labs(x = "Legal field", y = "Number of cases", title = "Cases by legal field")
```
```{r}
#Data Visualization 3
ggplot(cases, aes(x = `Professional Sanction`)) +
  geom_bar() +
  labs(x = "Professional sanction", y = "Number of cases", title = "Cases by professional sanction")
```
```{r}
library(dplyr)
library(lubridate)

cases <- cases %>%
  mutate(Month = floor_date(as.Date(Date), unit = "month"))

monthly_counts <- cases %>%
  count(Month, name = "n") 

MonthlyModel <- glm(n ~ Month, data = monthly_counts)
summary(MonthlyModel)

ggplot(monthly_counts, aes(x = Month, y = n)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Monthly Cases Over Time (Linear Regression)",
    x = "Month",
    y = "Number of Cases"
  ) +
  theme_minimal()

PMonthlyModel <- glm(n ~ Month, data = monthly_counts, family=poisson)
summary(PMonthlyModel)

monthly_counts$predicted <- predict(
  PMonthlyModel,
  newdata = monthly_counts,
  type = "response"
)

ggplot(monthly_counts, aes(x = Month, y = n)) +
  geom_point() +
  geom_line(
    aes(y = predicted),
    linewidth = 1
  ) +
  labs(
    title = "Monthly Cases Over Time (Poisson Regression)",
    x = "Month",
    y = "Number of Cases"
  ) +
  theme_minimal()


#Statistical Model
#model <- glm(n ~ Month, data = cases, family = poisson)
#summary(model)
#exp(coef(model))
```



