---
title: "Professional sanctions in AI hallucination cases"
format:
  html:
    embed-resources: true
    toc: true
---

Group members: Ishita Jain, Ahlam Negash, Abigail Minin

GitHub repository: [View our GitHub repository](https://github.com/ishitajain21/stat-learning-group-analysis)

Partner 1: Ishita Jain — research question, exploratory analysis, decision tree.

Partner 2: random forest, comparison, conclusions (name to be confirmed).

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 9, fig.height = 5.5, dpi = 144)
if (dir.exists("../.hw5-R-library")) .libPaths(c(normalizePath("../.hw5-R-library"), .libPaths()))
library(tidyverse)
library(rpart)
library(rpart.plot)
```

## Partner 1: Research question

Can party type, jurisdiction, AI-tool identification, and primary legal field
classify whether a documented AI hallucination case records a professional sanction?
The response is `Professional Sanction`, with Yes as the positive class.
This replaces the starter's question about counts over time to meet the
assignment's classification requirement.

The data come from Damien Charlotin's AI Hallucination Cases Database.
Each row represents a documented case, not a randomly sampled use of AI.
We predict the recorded label; No does not rule out later sanctions.

## Partner 1: Exploratory analysis

```{r}
cases <- read_csv("data/hallucination_cases.csv")
knitr::kable(cases |> count(`Professional Sanction`, name = "Cases"))
model_data <- cases |>
  filter(`Professional Sanction` %in% c("No", "Yes")) |>
  transmute(
    sanction = factor(`Professional Sanction`, levels = c("No", "Yes")),
    party = case_when(
      str_detect(str_to_lower(coalesce(`Party(ies)`, "")), "lawyer|prosecutor|defender") ~ "Lawyer involved",
      str_detect(str_to_lower(coalesce(`Party(ies)`, "")), "pro se") ~ "Pro se litigant",
      is.na(`Party(ies)`) ~ "Unknown",
      TRUE ~ "Other party"),
    jurisdiction = case_when(`State(s)` == "USA" ~ "USA",
                             is.na(`State(s)`) ~ "Unknown", TRUE ~ "Outside USA"),
    tool = case_when(str_to_lower(`AI Tool`) == "chatgpt" ~ "ChatGPT",
                    str_to_lower(`AI Tool`) %in% c("implied", "unidentified") | is.na(`AI Tool`) ~ "Unspecified or implied",
                    TRUE ~ "Other named tool"),
    legal_field = coalesce(`Legal Field Primary`, "Unknown"))
sanction_rate <- mean(model_data$sanction == "Yes")
ggplot(model_data, aes(sanction)) + geom_bar(fill = "steelblue") +
  labs(x = "Recorded professional sanction", y = "Cases",
       title = "Professional sanctions are uncommon") + theme_minimal()
party_summary <- model_data |> group_by(party) |>
  summarise(Cases = n(), Yes = sum(sanction == "Yes"),
            Sanction_rate = mean(sanction == "Yes"), .groups = "drop")
knitr::kable(party_summary, digits = 3)
ggplot(party_summary, aes(reorder(party, Sanction_rate), Sanction_rate)) +
  geom_col(fill = "darkorange") + coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Party category", y = "Proportion with sanction",
       title = "Recorded sanction rates by party type") + theme_minimal()
```

There are `r nrow(cases)` cases; `r nrow(cases) - nrow(model_data)` are
excluded for missing or unrecognized responses. Among retained cases,
`r round(100 * sanction_rate, 2)`% record a sanction. The imbalanced response
means high accuracy can conceal poor detection of Yes cases.
The party chart compares rates rather than counts to account for category
sizes. Lawyer involvement may be informative because professional discipline
applies to legal professionals. These associations do not establish causation.

## Shared modeling setup

```{r}
set.seed(154)
train_id <- sample(seq_len(nrow(model_data)), round(0.7 * nrow(model_data)))
train <- model_data[train_id, ]
test <- model_data[-train_id, ]
# Learn common fields from training data only; group unseen test levels.
common_fields <- train |> count(legal_field, sort = TRUE) |>
  slice_head(n = 8) |> pull(legal_field)
prepare_predictors <- function(d) {
  d |> mutate(
    legal_field = factor(ifelse(legal_field %in% common_fields, legal_field, "Other"),
                         levels = c(common_fields, "Other")),
    party = factor(party, levels = c("Lawyer involved", "Pro se litigant", "Other party", "Unknown")),
    jurisdiction = factor(jurisdiction, levels = c("USA", "Outside USA", "Unknown")),
    tool = factor(tool, levels = c("ChatGPT", "Unspecified or implied", "Other named tool")))
}
train <- prepare_predictors(train)
test <- prepare_predictors(test)
model_formula <- sanction ~ party + jurisdiction + tool + legal_field
knitr::kable(data.frame(Set = c("Training", "Test"),
                       Cases = c(nrow(train), nrow(test)),
                       Yes_rate = c(mean(train$sanction == "Yes"), mean(test$sanction == "Yes"))), digits = 3)
```

The seed-154 70/30 split gives `r nrow(train)` training and `r nrow(test)`
test cases. Partner 2 must reuse these processed sets and `model_formula`.
Outcome text, monetary penalties, and case narratives are excluded because
they could reveal the response; names and source links are also excluded.
This is a retrospective random holdout, not a forecast of future cases.

## Partner 1: Decision tree

```{r}
set.seed(154)
tree_full <- rpart(model_formula, data = train, method = "class",
                   parms = list(split = "gini"),
                   control = rpart.control(cp = 0.001, xval = 10))
printcp(tree_full)
best_cp <- tree_full$cptable[which.min(tree_full$cptable[, "xerror"]), "CP"]
tree_cv <- prune(tree_full, cp = best_cp)
# Prespecified exploratory tree; test data do not choose its settings.
tree_model <- tree_full
rpart.plot(tree_model, extra = 104, fallen.leaves = TRUE)
tree_model
leaf_count <- sum(tree_model$frame$var == "<leaf>")
root_variable <- tree_model$frame$var[1]
if (root_variable != "<leaf>") {
  root_paths <- path.rpart(tree_model, nodes = c(2, 3), print.it = FALSE)
  knitr::kable(data.frame(
    Rule = vapply(root_paths, function(x) tail(x, 1), character(1)),
    Cases = tree_model$frame[c("2", "3"), "n"],
    Yes_probability = tree_model$frame$yval2[match(c("2", "3"), rownames(tree_model$frame)), 5]), digits = 3)
}
leaf_id <- as.integer(rownames(tree_model$frame)[which(tree_model$frame$var == "<leaf>")[1]])
leaf_path <- path.rpart(tree_model, nodes = leaf_id, print.it = FALSE)[[1]]
leaf_row <- match(as.character(leaf_id), rownames(tree_model$frame))
leaf_class <- levels(train$sanction)[tree_model$frame$yval[leaf_row]]
leaf_probability <- tree_model$frame$yval2[leaf_row, 5]
```

The tree reduces Gini impurity to create more homogeneous sanction groups.
The exploratory tree uses a prespecified CP of 0.001 and has `r leaf_count`
terminal nodes. Training-only ten-fold cross-validation instead selects
CP = `r signif(best_cp, 4)`, yielding
`r sum(tree_cv$frame$var == "<leaf>")` terminal nodes. Its preference for a
single-node tree shows that these splits do not improve cross-validated
overall error enough to justify their complexity. We interpret the larger
tree as exploratory; the test set does not choose its settings.
`r if (root_variable == "<leaf>") "Cross-validation selected no splits; the tree predicts the majority class for every case." else paste0("The first split uses ", root_variable, ". The table shows its routing rules and the training sanction proportions on each side. The child with the larger Yes probability records sanctions more often.")`

One terminal node follows
`r if (length(leaf_path) > 1) paste(leaf_path[-1], collapse = "; then ") else "all cases at the root"`.
It predicts `r leaf_class`, with estimated Yes probability
`r round(leaf_probability, 3)`. This is a training-node proportion, not
certainty about an individual case.

```{r}
if (length(tree_full$variable.importance)) {
  vi <- sort(tree_full$variable.importance, decreasing = TRUE)
  knitr::kable(data.frame(Predictor = names(vi), Importance = as.numeric(vi)), digits = 2)
}
```

Importance values from the candidate tree measure split improvements,
including surrogate contributions. They have no sign and are not causal effects.

## Partner 1: Held-out performance

```{r}
classification_metrics <- function(prediction, actual) {
  cm <- table(Predicted = factor(prediction, levels = c("No", "Yes")),
              Actual = factor(actual, levels = c("No", "Yes")))
  sensitivity <- cm["Yes", "Yes"] / sum(cm[, "Yes"])
  specificity <- cm["No", "No"] / sum(cm[, "No"])
  c(Accuracy = sum(diag(cm)) / sum(cm), Error = 1 - sum(diag(cm)) / sum(cm),
    Sensitivity = sensitivity, Specificity = specificity,
    Balanced_accuracy = mean(c(sensitivity, specificity)))
}
tree_prob <- predict(tree_model, test, type = "prob")[, "Yes"]
tree_class <- factor(ifelse(tree_prob >= 0.5, "Yes", "No"), levels = c("No", "Yes"))
tree_cm <- table(Predicted = tree_class, Actual = test$sanction)
tree_cm
tree_metrics <- classification_metrics(tree_class, test$sanction)
baseline_metrics <- classification_metrics(rep("No", nrow(test)), test$sanction)
knitr::kable(rbind(Decision_tree = tree_metrics, Always_No = baseline_metrics), digits = 3)
```

At a 0.5 Yes-probability cutoff, test accuracy is
`r round(100 * tree_metrics["Accuracy"], 2)`%, versus
`r round(100 * baseline_metrics["Accuracy"], 2)`% for always predicting No.
Sensitivity is `r round(100 * tree_metrics["Sensitivity"], 2)`%: the proportion
of recorded Yes cases correctly identified. Specificity is
`r round(100 * tree_metrics["Specificity"], 2)`%, the proportion of recorded
No cases correctly identified. Balanced accuracy is
`r round(100 * tree_metrics["Balanced_accuracy"], 2)`%, giving both classes equal weight.

The cross-validation result favors the majority-class baseline over a
complex tree, so the exploratory splits should not be treated as a validated
improvement. The confusion matrix and sensitivity are essential because predicting No
can yield high accuracy. The tree gives readable rules, but rare positives
and small leaves can limit reliable detection. Reporting practices,
different jurisdictional rules, and later changes in outcomes limit these
results. Related cases or repeated parties can make a random holdout more
optimistic than prediction for entirely new parties. These results describe
this database, not sanction rates among all filings or causal effects of AI tools.

## Partner 2: Random forest

Pending Partner 2. Fit a random forest with `model_formula`, `train`, and
`test` above. Use Yes as the positive class and a 0.5 cutoff for the initial
comparison. Evaluate with `classification_metrics()`.

## Partner 2: Model comparison and conclusions

Pending Partner 2 and shared review. Compare with `tree_metrics` and the
always-No baseline using accuracy, sensitivity, specificity, balanced
accuracy, and interpretability. Discuss class imbalance and data limitations.
